---
title: "STA_135_Final_Project"
author: "Adam Hetherwick"
date: "2024-06-05"
output: html_document
---

## 

```{r}
library(HSAUR2)
data(USairpollution)

head(USairpollution)
```

## Data Description and Summary

Goal: Want to see which predictors are the best at predicting degree of air pollution in a city as measure by the sulfur dioxide content in the air. Can also rank the performance of the cities.

```{r}
data <- USairpollution

dim(data)
```
Ensuring normality of the predictors (qqplots and shapiro test) (2-3)

```{r}
normality_test_results <- lapply(data, shapiro.test)
p_value <- sapply(normality_test_results, function(x) x$p.value)
t(data.frame(
  p_value
))

data_t = data
data_t$SO2 = log(data$SO2)
data_t$temp = log(data$temp)
data_t$manu = log(data$manu)
data_t$popul = log(data$popul)

normality_test_results2 <- lapply(data_t, shapiro.test)
p_value <- round(sapply(normality_test_results2, function(x) x$p.value), 4)
t(data.frame(
  p_value
))
```

```{r}
library(ggplot2)
library(tidyr)
library(psych)

data_predictors <- data_t[1:41, 2:7]

buckets <- ifelse(data_t$SO2 > median(data_t$SO2), 'high_pol', 'low_pol')
buckets <- as.factor(buckets)
data_buckets <- data_predictors
data_buckets$buckets <- buckets

pairs.panels(data_predictors,
             gap = 0,
             bg = c("red", "blue")[as.factor(buckets)],
             pch = 21)

ggplot(data_t, aes(x = SO2, y = rep('', 41))) +
  geom_boxplot(width = 0.3) +
  geom_boxplot() +
  labs(title = "Sulfur Dioxide Distribution",
       x = "",
       y = "") + 
  theme(
    plot.margin = unit(c(1, 1, 1, 1), "cm"),  # Adjust plot margins
    aspect.ratio = 0.2  # Adjust aspect ratio to make the plot skinnier
  )
```

Covariance Matrices

```{r}
library(reshape2)
library(scales)

low_pol_data = data_buckets[data_buckets$buckets == 'low_pol', ]
high_pol_data = data_buckets[data_buckets$buckets == 'high_pol', ]

low_cov_matrix = round(cov(low_pol_data[, 1:6]), 2)
high_cov_matrix = round(cov(high_pol_data[, 1:6]), 2)

low_cov_matrix
high_cov_matrix
```
Checking PSD of covariance matrices:

```{r}
is_positive_semidefinite <- function(matrix) {
  eigenvalues <- eigen(matrix)$values
  return(all(eigenvalues >= -1e-8))
}

is_psd_high <- is_positive_semidefinite(high_cov_matrix)
is_psd_low <- is_positive_semidefinite(low_cov_matrix)

print(paste("High pollution matrix is positive semidefinite:", is_psd_high))
print(paste("Low pollution matrix is positive semidefinite:", is_psd_low))

high_cov_eig = round(eigen(high_cov_matrix)$values, 4)
low_cov_eig = round(eigen(low_cov_matrix)$values, 4)

t(data.frame(
  'High_Pollution_Covariance_Matrix_Eigenvalues' = high_cov_eig,
  'Low_Pollution_Covariance_Matrix_Eigenvalues' = low_cov_eig
))
```

Looking at predictors' relationship with the response.

```{r}
data_buckets_full <- data_t
data_buckets_full$buckets <- ifelse(data$SO2 > median(data$SO2), 'high_pol', 'low_pol')

melted_data <- melt(data_buckets_full, id.vars = c("SO2", "buckets"))

ggplot(melted_data, aes(x = value, y = SO2, color = buckets)) +
  geom_point() +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Predictors vs. SO2", 
       x = "Predictor Value", y = "SO2", color = "Pollution Level") +
  scale_color_manual(values = c("high_pol" = "red", "low_pol" = "blue"))
```

Checking the population covariance matrices (Box's M test 4-2)

```{r}
n1 = dim(low_pol_data)[1]
n2 = dim(high_pol_data)[1]

v1 = n1-1
v2 = n2-1

p = 6

c1_num = (((1/v1) + (1/v2) - (1/(v1+v2))) * (2*p**2 + 3*p - 1))
c1_den = 6*(p + 1)
c1 = c1_num / c1_den

s_x = cov(low_pol_data[, 1:6])
s_y = cov(high_pol_data[, 1:6])
s_p = (((n1-1)*s_x) + ((n2-1)*s_y)) / (n1 + n2 - 2)

M_num = ((det(s_x)**(v1/2))*(det(s_y)**(v2/2)))
M_den = (det(s_p)**((v1/2) + (v2/2)))
M = M_num / M_den

u1 = -2*(1-c1)*log(M)
u1

qchisq(0.95, 0.5*p*(p+1))
```
Reject the null if u1 > qchisq(0.95, 0.5p(p+1))

Fitting Quadratic Discriminant Analysis to the data

```{r}
library(MASS)
library(caret)

data_t_full <- data_t
data_t_full$buckets <- buckets

qda_model <- qda(buckets ~ temp + manu + popul + wind + precip + predays, data = data_t_full)
predictions <- predict(qda_model)$class

qda_class_rate = sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
qda_class_rate

confusionMatrix(predictions, as.factor(data_t_full$buckets))
```

Fitting a Linear Discriminant Analysis to each predictor

```{r}
lda_cols = colnames(data_t_full[, 2:7])
lda_vars = c()
lda_so2 = rep(data_t$SO2, 6)
lda_rates = c()
lda_values = c()
lda_buckets = c()

for (i in 2:7) {
  lda_model <- lda(buckets ~ data_t[, i], data = data_t_full)
  lda_predictions <- predict(lda_model)$class
  class_rate = sum(lda_predictions == data_t_full$buckets)/length(data_t_full$buckets)
  
  lda_buckets = append(lda_buckets, lda_predictions)
  lda_rates = append(lda_rates, class_rate)
  lda_vars = append(lda_vars, rep(colnames(data_t_full)[i], dim(data_t_full)[1]))
  lda_values = append(lda_values, data_t_full[, i])
}

data.frame('Predictor' = lda_cols,
           'Classification Rate' = lda_rates)

lda_melted_data <- data.frame(
  "SO2" = lda_so2,
  "buckets" = lda_buckets,
  "variable" = lda_vars,
  "value" = lda_values
)

ggplot(lda_melted_data, aes(x = value, y = SO2, color = buckets)) +
  geom_point() +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Predictors vs. SO2", 
       x = "Predictor Value", y = "SO2", color = "Pollution Level") +
  scale_color_manual(values = c("high_pol" = "red", "low_pol" = "blue"))
```

Fitting LDA to whole dataset

```{r}
lda_model <- lda(buckets ~ temp + manu + popul + wind + precip + predays, data = data_t_full)
predictions <- predict(lda_model)$class

lda_class_rate = sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
lda_class_rate

confusionMatrix(predictions, as.factor(data_t_full$buckets))
```

Running PCA on full data.

```{r}
cov_matrix = cov(data_t[, -1])
ev = eigen(cov_matrix)$values
eve = eigen((cov_matrix))$vectors
l1 = eve[1:6, 1] %*% t(data_t[, -1])
l2 = eve[1:6, 2] %*% t(data_t[, -1])
plot(-l1[1,1:dim(data_t[, -1])[1]], l2[1, 1:dim(data_t[, -1])[1]])
plot(sort(-l1[1, 1:dim(data_t[, -1])[1]]))

total_var <- sum(ev)
var_explained <- c()

for (i in seq_len(length(ev))) {
  var_explained = append(var_explained, round(sum(ev[1:i])/total_var, 6))
}

temp_df <- data.frame(x = 1:length(var_explained), y = var_explained)
ggplot(temp_df, aes(x, y)) +
  geom_point() +
  geom_line(color = "red") +
  geom_text(aes(label = y), vjust = -0.5, size = 3) +
  labs(title = "Principal Component Variance Explained", 
       x = "Eigenvalue", y = "Variance Explained") +
  theme_minimal()

pca_result <- prcomp(data_t[, -1], scale. = TRUE, center = TRUE, retx = TRUE)
pca_scores <- pca_result$x[, 1:2]

pca_df <- as.data.frame(pca_scores)
names(pca_df) <- c("PC1", "PC2")

merged_df <- cbind(pca_df, data_t_full$buckets)

ggplot(merged_df, aes(x = PC1, y = PC2, color=buckets)) +
  geom_point() +
  labs(title = "PCA Scatter Plot", 
       x = "Principal Component 1", y = "Principal Component 2", color = 'Polution Level') + 
  theme_minimal() + 
  scale_color_manual(values = c("high_pol" = "red", "low_pol" = "blue"))
```

Running QDA/LDA on PCA divided data

```{r}
pca_df$buckets <- data_t_full$buckets

qda_model_pca <- qda(buckets ~ PC1 + PC2, data = pca_df)
predictions <- predict(qda_model_pca)$class
qda_class_rate = sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
confusionMatrix(predictions, as.factor(data_t_full$buckets))

lda_model_pca <- lda(buckets ~ PC1 + PC2, data = pca_df)
predictions <- predict(lda_model_pca)$class
lda_class_rate = sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
confusionMatrix(predictions, as.factor(data_t_full$buckets))
```

Support Vector Machine model comparison

```{r}
library(e1071)

data_scaled <- data_t_full[, -1]
data_scaled[, 1:6] = scale(data_scaled[, 1:6])

# Linear kernel
svm_model <- svm(buckets ~ temp + manu + popul + wind + precip + predays, 
                 data = data_scaled, type = 'C-classification', kernel = 'linear')
predictions <- predict(svm_model, data_scaled)
svm_class_rate_linear =  sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
svm_model

# Radial kernel
svm_model <- svm(buckets ~ temp + manu + popul + wind + precip + predays, 
                 data = data_scaled, type = 'C-classification', kernel = 'radial')
predictions <- predict(svm_model, data_scaled)
svm_class_rate_radial =  sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
svm_model

# polynomial kernel
svm_model <- svm(buckets ~ temp + manu + popul + wind + precip + predays, 
                 data = data_scaled, type = 'C-classification', kernel = 'poly')
predictions <- predict(svm_model, data_scaled)
svm_class_rate_poly =  sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
svm_model

t(data.frame(
  'Kernel_Method' = c("Linear", "Radial", "Polynomial"),
  'Accuracy_Rate' = c(svm_class_rate_linear, svm_class_rate_radial, svm_class_rate_poly)
))
```

Random Forest model comparison

```{r}
library(randomForest)

rf_model <- randomForest(buckets ~ temp + manu + popul + wind + precip + predays, 
                         data = data_scaled, ntree = 100)
predictions <- predict(rf_model, data_scaled)
rf_class_rate =  sum(predictions == data_t_full$buckets)/dim(data_t_full)[1]
rf_class_rate
confusionMatrix(predictions, data_t_full$buckets)
```


